Metropolitan Statistical Area Evaluation

### 2023-02-15 src/demo_data_wrangling.Rmd
added copyright line for house price data
cleaning up data, added pseudocode

---
title: "Metropolitan Statistical Area Evaluation"
author: "Grace Bianchi"
date: "5/12/2022"
output:   html_document
---

This report provides an explanatory overview of demographic and real estate data for metropolitan statistical areas. The code wrangles and analyzes data collected from government agencies to create metrics to explore the relationships between criteria and compare across MSAs. Specifically, this data analysis includes population growth, household income, employment rates, house prices, and property taxes. Preliminary analysis of population change and size narrowed down potential areas which experienced population growth rate greater than 0.5% and have populations greater than 300,000 people. In addition, 28 MSAs were removed from analysis, as requested by the client, due to hyper-competitive real estate markets. In brief, there are 44 metropolitan statistical areas with potential for ZNE Capital aqcuire real estate.

@@

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(janitor) #clean_names
library(DT) # create data table
library(scales) 
library(lubridate)
library(gridExtra) 
library(ggrepel)
library(kableExtra)
library(googlesheets4)
library(tigris) # Census shape files
library(sf)
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
```

### Population Growth
All states were considered with the exception of CA, HI, and AK. Initial analysis of Metropolitan Statistical Areas (MSAs) relied on Census data of annual population estimates, in which the population had to be greater than 300,000 individuals for the purpose of attracting investors. Population change from 2020 to 2021 was calculated by $$\frac{2021 estimate- 2020 estimate}{2020 estimate} * 100$$
MSAs with a population growth > 0.5%, were selected to gather further data on. The 0.5% growth was determined.

@@

```{r population growth data wrangling}
# read in the data MSA population data from Census
msa_pop_2021_data <- read_csv(here("data","raw", "msa_2021_pop_data.csv")) %>% 
  clean_names() %>% # data wrangling to make column headers in lower_snake_case
  filter(lsad == "Metropolitan Statistical Area") %>% # select MSA cities
  select("cbsa", "name", "popestimate2020", "popestimate2021", "residual2020", "residual2021") %>% # select data

  separate(col = "name",
           into = c("MSA", "state"), # split city and state data
           sep = ", ") %>% 
  mutate(pop_percent_change = (popestimate2021-popestimate2020)/popestimate2020 * 100)  # add population % change column

# create object with viable MSAs 
potential_MSAs <- msa_pop_2021_data %>% 
  filter(pop_percent_change > 0.5 ) %>% 
  filter(popestimate2021 >= 300000) %>% 
  mutate(pop_percent_change20_21 = round(pop_percent_change, digits = 3)) %>% 
  filter (!state %in% c("CA", "HI", "AK")) %>% 
  #separate(col = "state", into = c("state1", "state2", "state3"),    sep = "-") %>% 
  filter(!cbsa %in% c("10580", "10900", "12420", "12940", "14260", "15980", "18580", "19100", "19660", "19740", "23540", "25420", "26420", "27260", "29820", "34940", "36260", "36740", "38060", "39100", "39300", "39340", "41620", "45220", "45300", "46060", "46140", "39900")) %>%  # remove MSAs as instructed by Owen 
  mutate(normalized_pop2021 = (popestimate2021-min(popestimate2021))/ (max(popestimate2021)-min(popestimate2021))) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021)

@@

```{r population growth msa map, eval=FALSE}
# adding column for levels of population growth
pop_growth_pos <- potential_MSAs %>% 
  mutate(pop_pos = case_when(
    pop_percent_change20_21 > 2 ~ "high",
    pop_percent_change20_21 > 1 ~ "medium",
    pop_percent_change20_21 > 0 ~ "low"
  ))

# MSA shapefiles from Census
cbsa <- core_based_statistical_areas(cb = TRUE) %>% 
  mutate(CBSAFP = as.numeric(CBSAFP)) %>% 
  inner_join(pop_growth_pos, by = c("CBSAFP" = "cbsa"))

@@

# map of U.S. with MSAs highlighted
ggplot() +
  geom_polygon(data = states,
               aes(x = long, y = lat,
                   group = group)) + 
  geom_sf (data = cbsa, aes(fill = pop_pos)) +
  labs(title = "MSAs with the highest population growth in 2020")
```

Figure 1. Population growth in Metropolitan Statistical Areas (n=44) in 2020. Highlighted areas have a population > 300,000 individuals and a population growth rate greater than 0.5%. The colors of the MSAs represent incremental growth rates, in which low(green) is 0.5-1% increase, medium(blue) is 1-2% increase, and high(pink) is 2-3.5% increase. Data: U.S. Census Bureau, Population Division.

```{r population change visual, fig.height = 7, fig.align='center', eval = FALSE}
ggplot(data = positive_MSA_pop_change, aes(x = pop_percent_change20_21, y = reorder(MSA, pop_percent_change20_21))) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5)

@@

```{r household income data wrangling}
MSA_income_data <- read_csv(here("data","raw", "2021_income_data.csv")) %>% 
  row_to_names(1) %>% 
  clean_names() %>% 
  mutate(median_income = as.numeric(estimate_households_median_income_dollars)) %>% 
  mutate(income_moe_USD = as.numeric(margin_of_error_households_median_income_dollars)) %>% 
  separate(col = "geography",
           into = c("id", 'cbsa'),
           sep = 9) %>% 
  mutate(cbsa = as.numeric(cbsa)) # change column class from chr to dbl

# join population growth data set to household median income
income_potential_MSAs <- MSA_income_data %>% # rename row names
  select(cbsa, median_income) %>% 
  inner_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, median_income, pop_percent_change20_21, popestimate2021, normalized_pop2021) #reorder columns

@@

# data wrangling, only selecting data for potential MSAs
MSA_employment <- unemployment_data %>% 
  filter(year >= 2019) %>% 
  mutate(cbsa = ifelse(cbsa == '76750', "38860", cbsa)) %>% # change south portland, ME to corresponding cbsa
  mutate(cbsa = as.numeric(cbsa)) %>% 
  full_join(income_potential_MSAs) %>% 
  drop_na() %>% 
  select(cbsa, MSA, year, month, employment_rate)

avg_emp_yoy <- MSA_employment %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  mutate(avg_emp_YoY = round(avg_emp_YoY, digits = 4)) %>% 
  drop_na()

MSA_emp_income <- income_potential_MSAs %>% 
  full_join(avg_emp_yoy) %>% 
  filter(year == "2021") %>% 
  select(!c(avg_employment_rate, year))

@@

```{r}
# adding portland employment rates from https://data.bls.gov/pdq/SurveyOutputServlet
portland_employment_data <- unemployment_data %>% 
  filter(year >= 2019) %>% 
  filter(cbsa == "76750") %>% 
   group_by(year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  mutate(avg_emp_YoY = round(avg_emp_YoY, digits = 4))

employment_data_jan_2021 <- 100-4.9
employment_data_jan_2022 <- 100- 3.6

yoy <- (employment_data_jan_2022 - employment_data_jan_2021)/ employment_data_jan_2021 *100

```{r employment visuals, fig.align = "center", fig.show="hold", fig.height= 7, fig.width=10, eval = FALSE}
# visual of MSA employment rates, shows major drop during covid
ggplot(MSA_employment_yoy, aes(x = year, y = employment_rate, color = MSA)) +
  geom_line() +
  xlim(2021, 2022) +
  theme(legend.position = "none") +

  labs(x = "Year",
       y = "Employment Rate",
      title = "Employment Rate Trend since 2016")

# year over year change in employment rate, split up by year
ggplot(MSA_employment_yoy, aes(x = employment_YoY, y = reorder(MSA, employment_YoY), fill = emp_pos)) +
  geom_col(position = "identity", color = "black", size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  facet_grid(~year) +
  labs(x = "Year-over_year change in employment rate",
       y = " ")

@@

```{r median rent}
# median rents from hud
MSA_median_rent2021_data <- read_csv(here("data", "raw", "2021MSA_medianrent.csv")) %>% 
  separate(col = cbsasub21,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = ifelse(cbsa == 'M6400', "38860", cbsa)) %>% 
  mutate(cbsa = as.numeric(cbsa)) %>% 
  mutate(median_rent2021 = as.numeric(rent50_1)) %>% 
  select(cbsa, median_rent2021) %>% 
  right_join(potential_MSAs)

MSA_median_rent2022_data <- read_csv(here("data", "raw", "2022MSA_median_rent.csv")) %>% 
  separate(col = cbsasub22,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = ifelse(unknown == '38860M', "38860", cbsa)) %>% 
  mutate(median_rent2022 = as.numeric(rent_50_1)) %>% 
  select(cbsa, median_rent2022) %>% 
  mutate(cbsa = as.numeric(cbsa)) %>% 
  right_join(MSA_median_rent2021_data)

potential_MSAs <- MSA_median_rent2022_data %>% 
  left_join(MSA_emp_income)

@@

Median housing prices by MSA were obtained from the Association of Realtors for 2021.

@@

```{r house price data}
metro_house_cost_data <- read_csv(here("data","raw", "metro_home_costs", "metro-home-prices-2022-11-10.csv")) %>% 
  row_to_names(3) %>% 
  clean_names() %>% 
  separate(col = "metropolitan_area",
           into = c("cbsa", 'metro_area'),
           sep = 5) %>% 
  mutate(cbsa = as.numeric(cbsa)) %>% 
  mutate(x2021 = as.numeric(x2021)) %>% 
  mutate(x2021 = x2021 * 1000) %>% 
  rename(house_price_2021 = x2021)

metro_house_prices <- metro_house_cost_data %>%
  select(cbsa, house_price_2021) %>% 
  right_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021, median_income, avg_emp_YoY, median_rent2021, median_rent2022, house_price_2021) %>% 
  mutate(house_price_2021 = ifelse(cbsa == "12260", "284900", house_price_2021)) %>% 
  mutate(house_price_2021 = ifelse(cbsa == "32580", "259000", house_price_2021))

MSAs <-metro_house_prices %>% 
  filter(popestimate2021 >= 500000) %>% 
  distinct(cbsa, .keep_all = TRUE) # remove duplicate portland row
```

@@

```{r, house price visual, fig.height = 7, fig.align='center', eval = FALSE}
ggplot(metro_house_prices, aes(x = houseprice_change21_22, y = reorder(MSA, houseprice_change21_22))) +
  geom_col(fill = "dodgerblue4") +
  geom_text(aes(label = houseprice_change21_22),
            size = 4,
            hujust = 1, nudge_x = -1,
            color = "white") +
  theme_minimal()  +
  labs(x = "% change",
       y = "",
       title = "Change in House Prices in 2021")

@@

```{r data table}
# table seperated by MSAs
datatable(MSAs, extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)

#gs4_create("updated_MSAs")
sheet_write(MSAs, ss = "https://docs.google.com/spreadsheets/d/1lmTpSDwVANxdAg5sW87Q8l_gtig7M0EjPPJ6J5_7dwg/edit#gid=0", sheet = "MSAs")
```

```{r MSAs seperate into cities}
#  split up MSAs into cities to collect utility, rent, and landlord data.
msa_seperated <- MSAs %>% 
  separate_rows(MSA, sep = "-") %>% 
  rename(city = MSA) %>% # separate MSA cities into their own columns
  filter(! city == "")

#gs4_create("updated_MSAs")
sheet_write(msa_seperated, ss = "https://docs.google.com/spreadsheets/d/1lmTpSDwVANxdAg5sW87Q8l_gtig7M0EjPPJ6J5_7dwg/edit#gid=0", sheet = "cities")

# table seperated by city
datatable(msa_seperated, extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)

>>>>

---
title: "Metropolitan Statistical Area Evaluation"
author: "Grace Bianchi"
date: "5/12/2022"
output:   
  html_document: 
  code_folding: hide
---

This report provides an explanatory overview of demographic and real estate data for metropolitan statistical areas. The code wrangles and analyzes data collected from government agencies to create metrics to explore the relationships between criteria and compare across MSAs. Specifically, this data analysis includes population growth, household income, employment rates, house prices, and property taxes. Preliminary analysis of population change and size narrowed down potential areas which experienced population growth rate greater than 0.5% and have populations greater than 300,000 people. In addition, 27 MSAs were removed from analysis, as requested by the client, due to hyper-competitive real estate markets. In brief, there are 29 metropolitan statistical areas with potential for ZNE Capital aqcuire real estate.

@@

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(janitor) #clean_names
library(DT) # create data table
# library(scales) 
library(lubridate)
library(gridExtra) 
library(ggrepel)
library(kableExtra)
library(googlesheets4) # output tables to google sheet
library(tigris) # imports Census shape files
library(sf)
options(tigris_class = "sf")
options(tigris_use_cache = TRUE)
```

### Population Growth
All states were considered with the exception of CA, HI, and AK. Initial analysis of Metropolitan Statistical Areas (MSAs) relied on Census data of annual population estimates, in which the population had to be greater than 500,000 individuals for the purpose of attracting investors. Population change from 2020 to 2021 was calculated by $$\frac{2021 estimate- 2020 estimate}{2020 estimate} * 100$$
MSAs with a population growth > 0.5%, were selected to gather further data on. The 0.5% growth was determined.

@@

```{r population growth data wrangling}
# read in the data MSA population data from Census
msa_pop_2021_data <- read_csv(here("data","raw", "msa_2021_pop_data.csv")) %>% 
  clean_names() %>% # data wrangling to make column headers in lower_snake_case
  filter(lsad == "Metropolitan Statistical Area") %>% # select MSA cities
  select("cbsa", "name", "popestimate2020", "popestimate2021", "residual2020", "residual2021") %>% # select data 
  separate(col = "name",
           into = c("MSA", "state"), # split city and state data
           sep = ", ") %>% 
  mutate(pop_percent_change = (popestimate2021-popestimate2020)/popestimate2020 * 100)  # calculate population % change

# create object with viable MSAs 
potential_MSAs <- msa_pop_2021_data %>% 
  filter(pop_percent_change > 0.5 ) %>% # select areas with population growth greater than .5%
  filter(popestimate2021 >= 500000) %>% 
  mutate(pop_percent_change20_21 = round(pop_percent_change, digits = 3)) %>% 
  filter (!state %in% c("CA", "HI", "AK")) %>% # remove areas within these states
  filter(!cbsa %in% c("10580", "10900", "12420", "12940", "14260", "15980", "18580", "19100", "19660", "19740", "23540", "25420", "26420", "27260", "29820", "34940", "36260", "36740", "38060", "39100", "39300", "39340", "41620", "45220", "45300", "46060", "46140", "39900")) %>%  # remove MSAs as instructed by client 
  select(cbsa, MSA, state, pop_percent_change20_21, popestimate2021)

@@

```{r population growth map, eval=FALSE}
## optional visualization
# read in MSA shapefiles from Census
cbsa <- core_based_statistical_areas(cb = TRUE) %>% 
  mutate(CBSAFP = as.numeric(CBSAFP)) %>% 
  inner_join(potential_MSAs, by = c("CBSAFP" = "cbsa")) # merge msa data with shape files

@@

# map of U.S. with MSAs highlighted
ggplot() +
  geom_polygon(data = states,
               aes(x = long, y = lat,
                   group = group)) + 
  geom_sf (data = cbsa, fill = "lightgreen") +
  labs(title = "Metro Areas included in this analysis", x = "", y = "")
```

Figure 1. Population growth in Metropolitan Statistical Areas (n=29) in 2020-2021. Highlighted areas have a population > 300,000 individuals and a population growth rate greater than 0.5%. The colors of the MSAs represent incremental growth rates, in which low(green) is 0.5-1% increase, medium(blue) is 1-2% increase, and high(pink) is 2-3.5% increase. Data: U.S. Census Bureau, Population Division.

```{r population change visual, fig.height = 7, fig.align='center', eval = FALSE}
# create plot of MSAs sorted grom highest population growth to lowest
ggplot(data = positive_MSA_pop_change, aes(x = pop_percent_change20_21, y = reorder(MSA, pop_percent_change20_21))) + 
  geom_point(size = 3, color = "darkgreen") + 
  theme_minimal(base_size = 12.5)

@@

```{r household income data wrangling}
MSA_income_data <- read_csv(here("data","raw", "2021_income_data.csv")) %>% 
  row_to_names(1) %>% 
  clean_names() %>% 
  mutate(median_income = as.numeric(estimate_households_median_income_dollars),
  income_moe_USD = as.numeric(margin_of_error_households_median_income_dollars)) %>% 
  separate(col = "geography",
           into = c("id", 'cbsa'),
           sep = 9) %>% 
  mutate(cbsa = as.numeric(cbsa)) %>% # change column class from chr to dbl
  inner_join(potential_MSAs)

# join household median income to existing msa dataframe
potential_MSAs <- potential_MSAs %>% add_column(med_income = MSA_income_data$median_income)

@@

# data wrangling, only selecting data for potential MSAs
MSA_employment <- unemployment_data %>% 
  filter(year >= 2018) %>% 
  mutate(cbsa = ifelse(cbsa == '76750', "38860", cbsa)) %>% # change south portland, ME to corresponding cbsa
  mutate(cbsa = as.numeric(cbsa)) %>% 
  right_join(potential_MSAs) %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_rate = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100, 
          avg_emp_rate = round(avg_emp_rate, digits = 4)) %>% 
  drop_na() %>% 
  mutate(emp_pos = case_when( 
    avg_emp_rate > 0 ~ "TRUE",
    avg_emp_rate < 1 ~ "FALSE"))

MSA_employ_rate <- MSA_employment %>% 
  filter(year == "2021")

potential_MSAs$employ_change <- MSA_employ_rate$avg_emp_rate

@@

```{r employment visuals, fig.align = "center", fig.show="hold", fig.height= 7, fig.width=10, eval = FALSE}
# visual of MSA employment rates, shows major drop during covid
ggplot(MSA_employment, aes(x = year, y = avg_employment_rate, color = MSA)) +
  geom_line() +
  xlim(2019, 2022) +
  theme(legend.position = "none") +
  labs(x = "Year",
       y = "Employment Rate",
      title = "Employment Rate Trend since COVID-19 Pandemic")

# year over year change in employment rate, split up by year
ggplot(MSA_employment, aes(x = avg_emp_rate, y = reorder(MSA, avg_emp_rate))) +
  geom_col(position = "identity", aes(fill = emp_pos), size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  facet_grid(~year) +
  labs(x = "Change in employment rate",
       y = " ")

@@

```{r median rent}
# data wrangling 2021 median rent
MSA_median_rent2021_data <- read_csv(here("data", "raw", "2021MSA_medianrent.csv")) %>% 
  separate(col = cbsasub21,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = ifelse(cbsa == 'M6400', "38860", cbsa),
        cbsa = as.numeric(cbsa) ,
        median_rent2021 = as.numeric(rent50_1)) %>% # select median rent for 1 bedroom apartments
  select(cbsa, median_rent2021) %>% 
  right_join(potential_MSAs)

# 2022 median rent
MSA_median_rent2022_data <- read_csv(here("data", "raw", "2022MSA_median_rent.csv")) %>% 
  separate(col = cbsasub22,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = ifelse(unknown == "38860M" & areaname22 == "Portland, ME HUD Metro FMR Area", "38860", cbsa),
         cbsa = as.numeric(cbsa),
         median_rent2022 = as.numeric(rent_50_1)) %>% 
  select(cbsa, median_rent2022) %>% 
  right_join(potential_MSAs)

potential_MSAs <- potential_MSAs %>% 
  add_column(med_rent2021 = MSA_median_rent2021_data$median_rent2021, # add rent data
             med_rent_2022 = MSA_median_rent2022_data$median_rent2022)

@@

Median housing prices by MSA were obtained from the Association of Realtors for 2021. 
Copyright ©[2021] NATIONAL ASSOCIATION OF REALTORS®. All rights reserved.  Reprinted with permission.

@@

```{r house price data}
metro_house_cost_data <- read_csv(here("data","raw", "metro_home_costs", "metro-home-prices-2022-11-10.csv")) %>% 
  row_to_names(3) %>% # make sure column headers are correct
  clean_names() %>% 
  separate(col = "metropolitan_area",
           into = c("cbsa", 'metro_area'),
           sep = 5) %>% 
  mutate(cbsa = as.numeric(cbsa),
        x2021 = as.numeric(x2021), 
        house_price_2021 = x2021 * 1000) %>% # data is proivded as thousands, so multiply to get correct ratio
  right_join(potential_MSAs) %>% 
  mutate(house_price_2021 = ifelse(cbsa == "12260", "284900", house_price_2021), 
        house_price_2021 = ifelse(cbsa == "32580", "259000", house_price_2021))

potential_MSAs$house_price_2021 <- metro_house_cost_data$house_price_2021

@@

```{r data table}
# table seperated by MSAs
datatable(potential_MSAs, extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)

#gs4_create("real_estate_inputs")
#output data into google sheet
sheet_write(potential_MSAs, ss = "https://docs.google.com/spreadsheets/d/1lmTpSDwVANxdAg5sW87Q8l_gtig7M0EjPPJ6J5_7dwg/edit#gid=0", sheet = "MSA_demographics")
```

```{r MSAs seperate into cities}
#  split up MSAs into cities to collect utility, rent, and landlord data.
msa_seperated <- potential_MSAs %>% 
  separate_rows(MSA, sep = "-") %>% 
  rename(city = MSA) %>% # separate MSA cities into their own columns
  filter(! city == "")

#gs4_create("real_estate_inputs")
sheet_write(msa_seperated, ss = "https://docs.google.com/spreadsheets/d/1lmTpSDwVANxdAg5sW87Q8l_gtig7M0EjPPJ6J5_7dwg/edit#gid=0", sheet = "cities")

@@
